---
title: "Unequal vaccine distribution leads to lopsided global recovery"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T,
  collapse = T,
  message = F,
  fig.width = 6, 
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "70%",
  dpi = 300,
  comment = "#>"
)
```

This is a documentation for the analysis behind a [story](https://www.thejakartapost.com/news/2021/08/01/slow-vaccination-takes-toll-on-indonesia-in-uneven-global-recovery.html){target="_blank"} on the July 2021 World Economic Outlook (WEO) from the International Monetary Fund (IMF).


## Packages

```{r Load the packages}
library(tidyverse)
library(data.table)
library(readxl)
library(lubridate)
library(glue)
library(WDI)
library(ggbeeswarm)
library(ggtext)
library(ggrepel)
```


## World Economic Outlook

We can get the WEO data from the IMF's [website](https://www.imf.org/en/Publications/WEO/Issues/2021/07/27/world-economic-outlook-update-july-2021#Data%20Tools){target="_blank"}. It is available in an Excel format. The file presents various indicators, but we will focus on the outlook for real gross domestic product (GDP) growth.

```{r Get WEO data}
weo_raw <- read_excel(
  "data/IMF_202107_WEO_raw.xlsx",
  sheet = "Real GDP Growth",
  na = "",
  skip = 4
)

weo_cleaned <- weo_raw %>% 
  select_if(function(x) {!all(is.na(x))}) %>% 
  rename(
    country = 1,
    actual_2019 = 2,
    actual_2020 = 3,
    forecast_2021 = 4,
    forecast_2022 = 5,
    revision_2021 = 6,
    revision_2022 = 7
  ) %>% 
  dplyr::filter(!is.na(country), !is.na(actual_2019)) %>% 
  mutate(
    country = str_remove_all(country, "2/"),
    country = str_replace(country, "Korea", "South Korea")
  )

weo_actual_outlook <- weo_cleaned %>% 
  pivot_longer(
    2:5,
    names_to = c("outlook", "year"),
    names_sep = "_",
    values_to = "growth",
    values_transform = list(growth = as.double)
  ) %>% 
  select(!starts_with("revision"))

weo_revision <- weo_cleaned %>% 
  select(country, starts_with("revision")) %>% 
  pivot_longer(
    2:3,
    names_to = c(NA, "year"),
    names_sep = "_",
    values_to = "revision"
  )

weo_tidy <- weo_actual_outlook %>% 
  left_join(weo_revision, by = c("country", "year"))

glimpse(weo_tidy)
```

There is an outlook for this year and 2022 in the data. But it contains only 120 observations because the IMF provides an outlook for selected countries.

The report came at a time when Indonesia was grappling with a resurgence in cases and deaths driven by the highly transmissible delta variant. So it will be interesting to see how the archipelagic country compares with its peers. We can check the bottom 10 countries in terms of outlook revision.

```{r How does Indonesia compare with other countries}
weo_tidy %>% 
  dplyr::filter(year == 2021) %>% 
  arrange(revision) %>% 
  head(10)
```

The Fund made the steepest downward revision at 3 percentage points for India between April and July. It was followed by Malaysia and Philippines. For Indonesia, the outlook was revised down by 0.4 percentage points. 

The variation in outlook revision suggests the symptoms of a divergent path in global economic recovery among countries, depending on not only whether they are battling rising cases and deaths but also how fast they are at inoculating their citizens. Hence, we will create a chart to show it.

As usual, we will first create a custom ggplot2 theme.

```{r Create custom ggplot theme}
theme_dfr <- function(..., base_size = 12, rel_size = 0.75) {
  
  theme(
    text = element_text(size = base_size, color = "#616161"),
    axis.title = element_blank(),
    axis.text = element_text(size = rel(rel_size)),
    axis.line.x = element_line(color = "black"),
    axis.ticks = element_blank(),
    panel.background = element_rect(fill = "white"),
    panel.grid.major.x  = element_blank(),
    panel.grid.major.y = element_line(color = "#E0E0E0"),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(
      face = "bold", 
      color = "black", 
      size = rel(1)
    ),
    plot.subtitle = element_text(
      size = rel(rel_size),
      margin = margin(b = 12.5)
    ),
    plot.title.position = "plot",
    plot.caption = element_text(
      size = rel(rel_size),
      hjust = 0,
      margin = margin(t = 12.5)
    ),
    plot.caption.position = "plot",
    ...
  )
  
}
```

To show the variation in outlook revision, we will use a lollipop chart. We will use different colors for countries that got an upward revision and those that got a downward revision to emphasize the divergence. And since there are dozens of countries, we will set the aspect ratio of our chart to `1` so it doesn't take up spaces.

```{r Plot WEO revision, fig.asp=1}
weo_2021_revision <- weo_tidy %>%
  dplyr::filter(year == "2021") %>% 
  mutate(country = fct_reorder(country, revision))


ggplot(weo_2021_revision, aes(x = revision, y = country)) +
  geom_vline(xintercept = 0, color = "#E68F7E") +
  geom_segment(
    aes(xend = 0, yend = country, color = revision >= 0),
    lwd = 1.5,
    alpha = 0.25,
    show.legend = F
  ) +
  geom_point(
    aes(fill = revision >= 0),
    pch = 21,
    color = "white",
    size = 2.5
  ) +
  scale_x_continuous(
    breaks = seq(-3, 3),
    limits = c(-3, 3),
    position = "top"
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#2477B3", "FALSE" = "#E66439"),
    labels = c("TRUE" = "Upward revision", "FALSE" = "Downward revision")
  ) +
  scale_color_manual(values = c("TRUE" = "#2477B3", "FALSE" = "#E66439")) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(
    title = "Indonesia lags behind in lopsided global recovery",
    subtitle = str_c(
      "Revision in 2021 real GDP growth outlook between April and July ", 
      "(percentage point)"
    ),
    x = NULL,
    y = NULL,
    caption = "Source: International Monetary Fund (IMF)\nChart: @dzulfiqarfr"
  ) + 
  theme_dfr() +
  theme(
    axis.text.y = element_text(hjust = 0, color = "#616161"),
    axis.line.x = element_blank(),
    panel.grid.major.x = element_line(color = "#E0E0E0"),
    legend.title = element_blank(),
    legend.text = element_text(size = rel(0.75), color = "#616161"),
    legend.position = c(0.06, 1.085),
    legend.key = element_blank(),
    legend.key.height = unit(0.25, "cm"),
    legend.key.width = unit(0.1, "cm"),
    legend.background = element_blank(),
    plot.subtitle = element_text(margin = margin(b = 25), color = "#616161")
  )
```


## Vaccination campaign

The IMF stated that the latest WEO showed a split among countries based on their vaccination rate. The higher the vacicnation rate the better outlook they have.

We can check how far the real GDP growth outlook varies with vaccination rate using vaccination data from [Our World in Data (OWID)](https://github.com/owid/covid-19-data/tree/master/public/data){target="_blank"}. We will take a look at the share of population partially vaccinated against COVID-19 at the latest date available. This metric shows the bare minimum in preventing people from contracting the virus.

```{r Get vaccination data}
url_owid <- "https://covid.ourworldindata.org/data/owid-covid-data.csv"

vax_raw <- fread(
  url_owid, 
  select = c("iso_code", "location", "date", "people_vaccinated_per_hundred")
)

# Remove observations by continent, which are denoted by "OWID" prefix in
# the `iso_code`
vax_latest <- vax_raw %>% 
  dplyr::filter(!str_detect(iso_code, "^OWID")) %>% 
  group_by(iso_code) %>% 
  dplyr::filter(!is.na(people_vaccinated_per_hundred)) %>%
  dplyr::filter(date == last(date)) %>% 
  ungroup()

glimpse(vax_latest)
```

Having imported the vaccination data, we can now join it with the WEO data.

```{r Join WEO and vaccination data}
# Call `parse_character()` to make sure values in `country` in WEO data 
# matches those in `location` in OWID data
weo_2021 <- weo_tidy %>% 
  dplyr::filter(year == 2021) %>% 
  mutate(country = parse_character(country))

weo_vax <- weo_2021 %>% 
  left_join(vax_latest, by = c("country" = "location"))

glimpse(weo_vax)
```

We can now create a scatter plot to see how real GDP growth outlook varies with vaccination rate. 

```{r Plot GDP outlook against vaccination rate}
today <- format(Sys.Date(), "%b %d, %Y")

ggplot(weo_vax, aes(people_vaccinated_per_hundred, growth)) +
  geom_point(
    pch = 21,
    fill = "black",
    color = "white",
    size = 2.5,
    alpha = 0.75
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    lwd = 1,
    lty = "dashed"
  ) +
  geom_text(
    aes(label = country),
    check_overlap = TRUE,
    size = 2.5,
    hjust = 0,
    nudge_x = 1,
    nudge_y = 0.25
  ) +
  scale_x_continuous(
    breaks = seq(0, 80, 20),
    limits = c(0, 80)
  ) +
  scale_y_continuous(
    breaks = seq(0, 10, 2),
    limits = c(0, 10),
    expand = c(0, 0),
    position = "right"
  ) +
  labs(
    title = "Better outlook for countries with higher vaccination rate",
    subtitle = str_c(
      "Real GDP growth outlook for 2021 & ",
      "share of population partially vaccinated against COVID-19*\n",
      "(percent)"
    ),
    x = "Share of population partially vaccinated*",
    y = "Real GDP growth outlook",
    caption = str_c(
      glue("*At latest date available as of {today}"),
      "\nSource: International Monetary Fund (IMF); Our World in Data",
      "\nChart: @dzulfiqarfr"
    )
  ) +
  theme_dfr() +
  theme(
    axis.title = element_text(size = rel(0.75)),
    panel.grid.major.x = element_line(color = "#E0E0E0")
  )
```

The scatter plot suggests a positive correlation between vaccination rate and real GDP growth outlook. It also shows that countries with higher vaccination rate tend to be richer countries, such as United Kingdom, Canada, France, Germany and the United States.

We can see such a trend more clearly by looking at vaccination rate across countries based on their income group. To group countries by their income level, we can use the income classification data from the World Bank using the **WDI** package.

```{r Get income classification}
wdi_country_raw <- as_tibble(WDI::WDI_data$country)

wdi_income_group <- wdi_country_raw %>% 
  select(iso3c, income)

glimpse(wdi_income_group)
```

We can now join the income classification data to the COVID-19 vaccination data. But not all countries in the vaccination data are available in the income classification data. We will drop countries with missing income classification.

```{r Join income classification and vaccination data}
vax_income_group <- vax_latest %>% 
  left_join(wdi_income_group, by = c("iso_code" = "iso3c")) %>%
  dplyr::filter(!is.na(income))

glimpse(vax_income_group)
```

We can now check the median vaccination rate by income group.

```{r Check average vaccination rate by income group}
vax_income_group_median <- vax_income_group %>% 
  group_by(income) %>% 
  summarize(median_vax_rate = median(people_vaccinated_per_hundred)) %>% 
  ungroup() %>% 
  arrange(desc(median_vax_rate))

vax_income_group_median
```

The median share of people who have received their first shot was `r vax_income_group_median[1, 2]` percent for high income countries and `r vax_income_group_median[4, 2]` percent for low income countries. This shows a very wide gap.

We can create one more chart to show the gap in vaccination rate by income group. We will plot vaccination rate by country with income group mapped to the x-axis. We will also show the median vaccination rate for each income group using a cross bar.

```{r Plot vaccination rate by income group}
annotation_country <- vax_income_group %>% 
  dplyr::filter(
    location %in% c(
      "Nepal",
      "Mongolia",
      "Nauru",
      "Gibraltar"
    )
  )

annotation_indonesia <- vax_income_group %>% 
  dplyr::filter(location == "Indonesia")

vax_income_group_sorted <- vax_income_group %>% 
  mutate(income = fct_reorder(income, people_vaccinated_per_hundred))

ggplot(vax_income_group_sorted, aes(income, people_vaccinated_per_hundred)) +
  geom_quasirandom(
    aes(fill = iso_code == "IDN"),
    pch = 21,
    color = "white",
    size = 2.25,
    show.legend = F,
    alpha = 0.9
  ) +
  stat_summary(
    geom = "crossbar",
    fun = median,
    lwd = 0.5,
    color = "#2477B3"
  ) +
  geom_richtext(
    data = annotation_country,
    aes(x = income, y = people_vaccinated_per_hundred, label = location),
    size = 2.75,
    hjust = 0,
    nudge_x = 0.075,
    fill = "white",
    label.color = NA,
    label.padding = grid::unit(rep(0, 4), "pt")
  ) +
  geom_richtext(
    data = annotation_indonesia,
    aes(income, people_vaccinated_per_hundred, label = location),
    size = 2.75,
    hjust = 1,
    vjust = 0,
    nudge_x = -0.1,
    fill = "white",
    color = "#E66439",
    fontface = "bold",
    label.color = NA,
    label.padding = grid::unit(rep(0, 4), "pt")
  ) +
  geom_text_repel(
    data = tibble(x = "Low income", y = 1.22, label = "Median"),
    aes(x, y, label = label),
    size = 2.75,
    color = "#2477B3",
    fontface = "bold",
    nudge_x = -0.25,
    nudge_y = 8,
    segment.color = "#2477B3",
    segment.curvature = -0.1,
    segment.ncp = 5,
    segment.size = 0.75
  ) +
  scale_y_continuous(
    breaks = seq(0, 120, 30),
    limits = c(0, 120),
    expand = c(0, 0),
    position = "right"
  ) +
  scale_x_discrete(
    labels  = c(
      "Low income",
      "Lower-middle income",
      "Upper-middle income",
      "High income"
    )
  ) +
  scale_fill_manual(values = c("FALSE" = "#55CBF2", "TRUE" = "#E66439")) +
  labs(
    title = "Vaccine rollout is much faster in richer countries",
    subtitle = str_c(
      "Share of population vaccinated against COVID-19* ", 
      "by income group (percent)"
    ),
    x = NULL,
    y = NULL,
    caption = str_c(
      glue("*At latest date available as of {today}"),
      "\nSource: Our World in Data; World Bank",
      "\nChart: @dzulfiqarfr"
    )
  ) +
  theme_dfr() +
  theme(axis.ticks.x = element_line(color = "black"))
```